---
title: "LanguageFinderSandbox"
author: "Catherine Brockway"
date: "`r Sys.Date()`"
output: 
  rmdformats::robobook: 
    code_folding: hide    
    code_download: true
---

```{r environment}
library(reactable)
# https://glin.github.io/reactable/articles/examples.html#basic-usage 
```

# Landing Page

Find out which languages other than English are spoken in particular places.
Translations and interpretation services are key tools to increase accessibility to information. Access to information is especially important in relation to:

1) Disaster warnings and disaster response;
2) Healthcare, especially treatment options and recovery instructions;
3) Local government engagement;
4) Environmental justice; 
5) Workplace safety

See how many people speak a particular language in different places.

1) Understanding how speakers of a certain language are distributed across the U.S. can help organizations find collaborative partners in creating translations and interpretive services for that language.

2) Organizations that serve particular language communities can use this information to identify key service areas.

This project seeks to make data about languages spoken in the US accessible to decision-makers in social sector enterprises. Key variables for decision-makers include the detailed language spoken, ability to speak English, and potentially age, for organizations that serve specific age groups, e.g.: voting age population, K-12 children, seniors, etc. 


# State Data
```{r goals, eval=FALSE, include=FALSE}
# We want to create a shiny app that user can input criteria and output is a table with only the data they have asked for

# Must haves:
# Clean, clear table output
# Ability to select levels of all variables as input
# Sort all columns in output
# Ability to download data table as .csv

# Nice to have:
# Ability to choose which columns to include in table
# Grouped output (e.g. # of all speakers meeting criteria, then number of each subset of criteria)
```

```{r load csv with all variables}

lang_totals <- read.csv("LangAppData.csv")

```


```{r reactable all languages}
# this is an example table of what the output could look like
reactable(
lang_totals,
groupBy = c("LANP_label", "ST_label", "AgeGroup"),
searchable = TRUE,
showPageSizeOptions = TRUE,
striped = TRUE,
highlight = TRUE,
fullWidth = FALSE,
columns = list(
LANP_label = colDef(name = "Language", filterable = TRUE),
ENG_label = colDef(name = "Ability to Speak English", filterable = TRUE),
AgeGroup = colDef(name = "Age Group", filterable = TRUE),
ST_label = colDef(name = "State", filterable = TRUE),
n = colDef(name = "Number of Speakers", aggregate = "sum")
)
)
```

# County Data

# Census Tract Data
```{r census tract goals, eval=FALSE, include=FALSE}
# Ability to select census tracts on map to create table (since no one knows the names/codes of census tracts). 
```

# About the data

The data comes from the [US Census Bureau](https://www.census.gov/) [American Community Survey](https://www.census.gov/programs-surveys/acs), 5-year estimates from 2017-2021. This is the most reliable and most recent data available on languages spoken at home in the US. This data is collected using the following questions:

![Image of Language Spoken at Home question on the American Community Survey](language_600_q14.avif){width=50%}

$~$

The US Census Bureau reports 130 language categories, including 105 individual languages (e.g.: Spanish, Hawaiian) and 25 aggregated language categories (e.g. Other Malayo-Polynesian languages, Aleut languages. 

Where age information is given, ages were divided into three categories: Youth (age 5-17), Working Age (18-64), and Senior (65-99). Data for individuals under age 5 and above age 99 are not provided by the US Census Bureau. 

Where Ability to Speak English is given, this data is self-reported in part C of the question shown above.
