---
title: "PUMSLanguageFinder"
author: Catherine Brockway, PhD; UN CIFAL Honolulu; catherine.lee.brockway@gmail.com;Epifania Petelo
date: "`r Sys.Date()`"
output: 
  html_document:
    code_download: TRUE
    theme: sandstone
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r env setup, include=FALSE}
library(tidyverse)
library(tidycensus)
library(kableExtra)
library(reactable)
# https://glin.github.io/reactable/articles/examples.html#basic-usage 
library(scales)
library(rmdformats)
Sys.getenv("CENSUS_API_KEY")

options(reactable.theme = reactableTheme(
color = "hsl(100, 0%, 0%)",
backgroundColor = "hsl(0, 0%, 100%)",
borderColor = "hsl(233, 9%, 22%)",
stripedColor = "hsl(240, 3%, 94%)",
highlightColor = "hsl(95, 80%, 70%)",
inputStyle = list(backgroundColor = "hsl(0, 0%, 100%)"),
selectStyle = list(backgroundColor = "hsl(0, 0%, 100%)"),
pageButtonHoverStyle = list(backgroundColor = "hsl(0, 0%, 100%)"),
pageButtonActiveStyle = list(backgroundColor = "hsl(0, 0%, 100%)")
))
#set your census api key using instructions at https://walker-data.com/tidycensus/reference/census_api_key.html

```


Find out which languages other than English are spoken in particular places.
Translations and interpretation services are key tools to increase accessibility to information. Access to information is especially important in relation to:

1) Disaster warnings and disaster response;
2) Healthcare, especially treatment options and recovery instructions;
3) Local government engagement;
4) Environmental justice; 
5) Workplace safety

See how many people speak a particular language in different places.

1) Understanding how speakers of a certain language are distributed across the U.S. can help organizations find collaborative partners in creating translations and interpretive services for that language.

2) Organizations that serve particular language communities can use this information to identify key service areas.

This project seeks to make data about languages spoken in the US accessible to decision-makers in social sector enterprises. Key variables for decision-makers include the detailed language spoken, ability to speak English, and potentially age, for organizations that serve specific age groups, e.g.: voting age population, K-12 children, seniors, etc. 

# About the data

The data comes from the [US Census Bureau](https://www.census.gov/) [American Community Survey](https://www.census.gov/programs-surveys/acs), 5-year estimates from 2017-2021. This is the most reliable and most recent data available on languages spoken at home in the US. This data is collected using the following questions:

![Image of Language Spoken at Home question on the American Community Survey](language_600_q14.avif){width=50%}

$~$

The US Census Bureau reports 130 language categories, including 105 individual languages (e.g.: Spanish, Hawaiian) and 25 aggregated language categories (e.g. Other Malayo-Polynesian languages, Aleut languages. 

Where age information is given, ages were divided into three categories: Youth (age 5-17), Working Age (18-64), and Senior (65-99). Data for individuals under age 5 and above age 99 are not provided by the US Census Bureau. 

Where Ability to Speak English is given, this data is self-reported in part C of the question shown above.

# State Level Language Data

```{r get US data, eval=FALSE, include=FALSE}
lang_pums_us <- get_pums(
  variables = c("LANP", "AGEP", "ENG"),
  state = "all",
  survey = "acs5",
  year = 2022,
  recode = TRUE
)

```

```{r eval=FALSE, include=FALSE}
nonEng_us <- lang_pums_us %>%
  filter(LANP != "000N")%>%
  select(PWGTP, AGEP, ST_label, ENG_label, LANP_label)

nonEng_us$ST_label <- as.factor(nonEng_us$ST_label)
nonEng_us$ENG_label <- as.factor(nonEng_us$ENG_label)
nonEng_us$ENG_label <- factor(nonEng_us$ENG_label, levels = c('Not at all', 'Not well', 'Well', 'Very well'))
nonEng_us$LANP_label <- as.factor(nonEng_us$LANP_label)

nonEng_us <- within(nonEng_us, {   
  AgeGroup <- NA # need to initialize variable
  AgeGroup[AGEP < 18] <- "Youth"
  AgeGroup[AGEP >= 18 & AGEP < 65] <- "Adult"
  AgeGroup[AGEP >= 65] <- "Senior"
   } )

nonEng_us$AgeGroup <- as.factor(nonEng_us$AgeGroup)
nonEng_us$AgeGroup <- factor(nonEng_us$AgeGroup, levels = c('Youth', 'Adult', 'Senior'))

totals_us <- nonEng_us%>%
  count(LANP_label, wt = PWGTP)%>%
  arrange(desc(n))
```


```{r reactable US languages, eval=FALSE, include=FALSE}
reactable(
totals,
searchable = TRUE,
showPageSizeOptions = TRUE,
striped = TRUE,
highlight = TRUE,
columns = list(
LANP_label = colDef(name = "Language"),
n = colDef(name = "Number of Speakers")
)
)
```


# Languages in Public Use Microdata Areas

If you work within a smaller geography than a state, you can use the tables below to find data for [Public Use Microdata Areas](https://www.census.gov/programs-surveys/geography/guidance/geo-areas/pumas.html), or PUMAs, which are the smallest geography available for detailed language used at home. Each PUMA contains a population of at least 100,000 people. Follow the instructions below to find the PUMA that best fits your area of interest. 

## Find your PUMA (Public Use Microdata Area)

Visit this interactive map tool to find the PUMA(s) you want to look at: [PUMA 2020 Boundaries](https://www.arcgis.com/home/item.html?id=da5ec62f0b5d49a2aecaeb14d6940fe1). Click the blue button on the right side of the screen: "Open in Map Viewer". Zoom in to your area of interest and click on the area that most closely matches the geography you want to focus on. The 5 digit code next to "PUMA" is the code you will use in the tables below. "Name" is the name of the PUMA you can use in your reporting. 



# Texas
```{r get data, include=FALSE}
st <- "TX"

lang_pums <- get_pums(
  variables = c("PUMA20", "LANP", "AGEP", "ENG"),
  state = st,
  survey = "acs5",
  year = 2022,
  recode = TRUE
)

nonEng <- lang_pums %>%
  filter(LANP != "000N")%>%
  select(PWGTP, AGEP, PUMA20, ENG_label, LANP_label)


nonEng$ENG_label <- as.factor(nonEng$ENG_label)
nonEng$ENG_label <- factor(nonEng$ENG_label, levels = c('Not at all', 'Not well', 'Well', 'Very well'))
nonEng$LANP_label <- as.factor(nonEng$LANP_label)

nonEng <- within(nonEng, {   
  AgeGroup <- NA # need to initialize variable
  AgeGroup[AGEP < 18] <- "Youth"
  AgeGroup[AGEP >= 18 & AGEP < 65] <- "Adult"
  AgeGroup[AGEP >= 65] <- "Senior"
   } )

nonEng$AgeGroup <- as.factor(nonEng$AgeGroup)
nonEng$AgeGroup <- factor(nonEng$AgeGroup, levels = c('Youth', 'Adult', 'Senior'))

lang_totals <- nonEng %>%
  count(LANP_label, ENG_label, AgeGroup, PUMA20, wt = PWGTP)%>%
  arrange(desc(n))
```

```{r reactable TX languages}

reactable(
lang_totals,
groupBy = c("LANP_label", "PUMA20", "AgeGroup"),
searchable = TRUE,
showPageSizeOptions = TRUE,
striped = TRUE,
highlight = TRUE,
fullWidth = FALSE,
columns = list(
PUMA20 = colDef(name = "PUMA", filterable = TRUE),
LANP_label = colDef(name = "Language", filterable = TRUE),
ENG_label = colDef(name = "Ability to Speak English", filterable = TRUE),
AgeGroup = colDef(name = "Age Group", filterable = TRUE),
n = colDef(name = "Number of Speakers", aggregate = "sum")
)
)

```

# Hawaiʻi
```{r get data, include=FALSE}
st <- "HI"

lang_pums <- get_pums(
  variables = c("PUMA20", "LANP", "AGEP", "ENG"),
  state = st,
  survey = "acs5",
  year = 2022,
  recode = TRUE
)

nonEng <- lang_pums %>%
  filter(LANP != "000N")%>%
  select(PWGTP, AGEP, PUMA20, ENG_label, LANP_label)


nonEng$ENG_label <- as.factor(nonEng$ENG_label)
nonEng$ENG_label <- factor(nonEng$ENG_label, levels = c('Not at all', 'Not well', 'Well', 'Very well'))
nonEng$LANP_label <- as.factor(nonEng$LANP_label)

nonEng <- within(nonEng, {   
  AgeGroup <- NA # need to initialize variable
  AgeGroup[AGEP < 18] <- "Youth"
  AgeGroup[AGEP >= 18 & AGEP < 65] <- "Adult"
  AgeGroup[AGEP >= 65] <- "Senior"
   } )

nonEng$AgeGroup <- as.factor(nonEng$AgeGroup)
nonEng$AgeGroup <- factor(nonEng$AgeGroup, levels = c('Youth', 'Adult', 'Senior'))

lang_totals <- nonEng %>%
  count(LANP_label, ENG_label, AgeGroup, PUMA20, wt = PWGTP)%>%
  arrange(desc(n))
```

```{r reactable HI languages}

reactable(
lang_totals,
groupBy = c("PUMA20", "LANP_label", "ENG_label"),
searchable = TRUE,
showPageSizeOptions = TRUE,
striped = TRUE,
highlight = TRUE,
fullWidth = FALSE,
columns = list(
PUMA20 = colDef(name = "PUMA", filterable = TRUE),
LANP_label = colDef(name = "Language", filterable = TRUE),
ENG_label = colDef(name = "Ability to Speak English", filterable = TRUE),
AgeGroup = colDef(name = "Age Group", filterable = TRUE),
n = colDef(name = "Number of Speakers", aggregate = "sum")
)
)

```
